---
title: "llm4np: language modeling for neuropsychology reports"
shorttitle: "AI for NP"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{llm4np: language modeling for neuropsychology reports}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

We use the [wizrd](https://github.com/lawremi/wizrd) package to produce
tooling for condensation and ontological tagging of neuropsychology reports.

## Using gpt-4o-mini to condense an example report

For this code to run without modifications, environment
variable `OPENAI_API_KEY` must have a valid setting.

```{r setup, message=FALSE}
stopifnot(nchar(Sys.getenv("OPENAI_API_KEY"))>0)
library(wizrd)
library(llm4np)
pretty_rmd <- function(x) {
    x |> strwrap() |> paste(">", text = _, collapse = "\n") |> cat()
}
```

We have retrieved an example PDF of a neuropsychological (NP) evaluation from 
[squarespace.com](https://static1.squarespace.com/static/50a3e393e4b07025e1a4f0d0/t/510b1429e4b0f6b4fb681381/1359680553898/de-identified+report+1.pdf).

**Note:** The `parse_nppdf` may emit warnings or errors during the table parsing
phase.  These refer to non-existent files, and seem to be false positives.  They
should be ignored.

`parse_nppdf` uses `pdftools` and `tabulapdf` to extract text and tables
from reports.
```{r doparse, message=FALSE}
deidpath = system.file("pdfs", "de-identified+report+1.pdf", package="llm4np")
deid = parse_nppdf(deidpath)
deid
```

We paste the report text into a single character vector for
convenient processing.
```{r dopa}
nt = gsub("\\n", " ", paste(deid@text, collapse=" "))
substr(nt,1130,1210)
```

We also have a view of an autism ontology based on a [2015 paper](https://pmc.ncbi.nlm.nih.gov/articles/PMC4000751/)
by Alexa McCray and colleages.
```{r lkpty}
data(nmdf)
head(nmdf)
```

Now we create an agent that will be prompted to use the ontology in a summary
of the NP evaluation.
```{r chkag}
ag = openai_agent("gpt-4o-mini") |> 
   instruct("You are a helpful assistant with training in neuropsychology.")
ag
```

We now use a templated prompt, using the report text and the ontology
in a summary.
```{r lksumm, results="asis"}
prompted_agent <- ag |> prompt_as("Please summarize the text {txt} in five short paragraphs 
    of at most 45 words.  Use the concepts in the 'val' field of {df} as much as possible, 
    choosing the most specific term that is relevant, and when you do use these terms, please 
    append the associated value of the 'tag' field in parentheses, prefixing the tag with 
    'PTO:'.  After producing the paragraphs, please produce three lists of tags that were 
    used, corresponding to 'strengths', 'weaknesses' and 'needs'.  Do not use headings with 
    the lists, just provide the lists as sentences preceded by the category.") 
prompted_agent |> predict(list(txt=nt, df=nmdf)) |> pretty_rmd()
```

## A second example


```{r lk2, results="asis"}
pecspa = system.file("pdfs", "PECS-Example-ADHD-Report.pdf", package="llm4np")
pecs = parse_nppdf(pecspa)
txt2 = gsub("\\n", " ", paste(pecs@text, collapse=" "))  # concatenate text
prompted_agent |> predict(list(txt=txt2, df=nmdf)) |> pretty_rmd()
```
